A general framework for learning planning action models

### 2018-07-09 doc/aij18/background.tex
small updates to background section aij18 paper

This section serves two purposes: first we introduce the basic planning concepts as well as the classical planning model that we will use throughout the rest of the paper; and secondly, we summarize the existing approaches to learn action models in classical planning and highlight our contributions via the related work.

@@

Like in PDDL~\cite{fox2003pddl2}, we assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ such that $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

A {\em state} $s$ is a full assignment of values to fluents, i.e. $|s|=|F|$, so the size of the state space is $2^{|F|}$. We explicitly include negative literals $\neg f$ in states since our approach handles with partial observability, thus adopting the \emph{open world assumption}.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $s=\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$ . A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan is {\em optimal} if it has minimum length.

@@

With regard to the observed states in a plan trace, we say that $\tup{s_0, s_1, \ldots, s_n}$ is a fully-observable (\FO) state trajectory if every state $s_i$ is a full assignment of values to fluents and the minimal action sequence to transit from state $s_i$ to state $s_{i+1}$ is composed of a single action; that is, $\theta(s_i,\tup{a})=s_{i+1}$. Otherwise, it is a partially-observable (\PO) state trajectory, meaning that at least one state $s_i$ is a partial assignment of values to fluents in which one or more literals are missing (formally, $|s_i|<|F|$). This implies that in a \PO state trajectory all fluents of a state $s_i$ may be missing, in which case, $s_i$ is a \emph{missing} or \emph{empty} state. This general definition of \PO gives rise to two particular cases:

\begin{itemize}
\item when \textbf{all} the $n-1$ intermediate states of a trajectory $s$ are \textbf{missing}, $s$ is a \emph{non-observable} (\NO) state trajectory
\item when \textbf{none} of the $n-1$ intermediate states of a trajectory $s$ is \textbf{missing}, we will refer to $s$ as a \POstar state trajectory.
\end{itemize}

@@

\begin{table}[hbt!]
\centering
\begin{tabular}{c|c|c|}
	     & \# intermediate states & state type \\ \hline
    \FO & $n-1$  & {\small $\forall i, 1 \leq i < n$}  \\  & & $s_i$ is a full assignment \\ \hline
    \multirow{1}{*}{\POstar} & $n-1$ & {\small $\exists i, 1 \leq i < n$}  \\ & & $s_i$ is a partial assignment \\ \hline
    \multirow{1}{*}{\PO} & $\leq n-1$ & {\small $\exists i, 1 \leq i < n$}   \\  & & $s_i$ is a partial assignment \\ \hline
    \NO & 0 & {\small $\forall i, 1 \leq i < n$}  \\  & & $|s_i|=0$  \\\hline
\end{tabular}
\caption{Classification of state trajectories accordingly to observation}
\label{tab:state_trajectory}
\end{table}

With regard to the observed actions in a plan trace, we say that an action sequence, $\tup{a_1, \ldots, a_n}$ is a \FO action sequence if it contains all the necessary actions to transit every state $s_{i-1}$ to the corresponding successor state $s_{i}$, for each {\small $1\leq i\le n$}. We say it is a \PO action sequence if at least one of these necessary actions is missing in the sequence and a \NO action sequence in case the sequence of observed actions is empty.

Given a  plan trace $\tau = \langle s_0, a_1, s_1, a_2, s_2, \ldots, a_n, s_n \rangle$ for a planning frame $\Phi=\tup{F,A}$, it holds that $a_i\in A$ for every action in $\tau$ and that $s_i \in \mathcal{L}(F)$ for each {\small $1\leq i\le n$}. Plan traces can be classified accordingly to the type of observed state trajectory (\FO, \POstar, \PO or \NO) and action sequence (\FO, \PO or \NO).

@@

The first column of Table \ref{table:models_comparison1} shows the constraints imposed on the input plan traces with regards to observability. Since all approaches except ours deal only with \FO action sequences, constraints are exclusively concerned with the type of state trajectory. This directly affects the complexity of the task, which can be sorted from the least to the most constrained following this order: 1) \NO, 2) \PO, 3) \POstar, and 4) \FO. Note that \PO is less constrained than \POstar because it also includes the possibility of having some missing state in the trajectory.

The task of learning from less constrained traces subsumes learning from more constrained ones. Consequently, approaches to learning from, for instance, traces with \PO state trajectories will also be able to learn from traces with \POstar state trajectories. All the approaches analyzed in this work accept the more constrained definition of partial observations of intermediate states \POstar, and most of them also allow the sequence of intermediate states to be empty. Exceptionally, a \NO state sequence in \LOCM is a fully-empty trajectory, with neither initial or final state.

@@

Most approaches assume that a set of predicates and a set of action headers are provided alongside the input traces. Others do not explicitly say so but the fact is that the predicates and the actions headers are easily extractable from the state sequence and action sequence of the plan traces, respectively. A requirement, however, for these two sets to be representative is that set of input plan traces comprises at least a grounded sample of all predicates and operator schemas of the domain model.

@@

Table \ref{table:models_comparison2} summarizes the main characteristics of the evaluation of the learned action models based on the type of evaluation method (first column of Table \ref{table:models_comparison2} -- almost all approaches rely on a comparison between the learned model and a ground-truth model), the metrics used in the comparison (second column of Table \ref{table:models_comparison2}) and the number of tested domains alongside the size of the training dataset (third column of Table \ref{table:models_comparison2}).

@@

\begin{table}
	\small
	\centering
	\begin{tabular}{ l | c | c | c }
		%& \multicolumn{1}{c}{Evaluation task}
        & \textbf{Evaluation method} & \textbf{Metrics} & \textbf{\#tested domains/}   \\
        &   &   & \textbf{training data size} \\
		\hline			
		\multirow{1}{*}{\ARMS} & cross-validation with a test set & error counting of \#$\pre$ satisfaction  & 6  \\
        & of plan traces & and redundancy & 1,600-4,320 actions\\ & & & (160 plan traces) \\
        \hline
         \SLAF &  manual checking wrt GTM &  ---   & 4\\ & & & 1,000 actions\\
         \hline
		\multirow{1}{*}{\LAMP} & checking wrt GTM  & error counting of extra & 4\\ & & and missing \#$\pre$ and \#$\eff$ & 1,300-6,100 actions\\
           & & & (100-200 plan traces) \\
         \hline
         \AMAN & checking wrt GTM &  error counting of extra &  3 \\ & & and missing \#$\pre$ and \#$\eff$ & 40-200 plan traces\\
         \hline
         \NOISTA & checking wrt GTM  & error counting of extra & 5\\ & & and missing \#$\pre$ and \#$\eff$ & 5,000-20,000 actions\\
         \hline
         \CAMA & checking wrt GTM  &  error counting of extra & 3 \\ & & and missing \#$\pre$ and \#$\eff$ & 15-75 plan traces\\
         \hline
		\LOCMtwo & manual checking wrt GTM &  ---  &   --- \\
         \hline
		\FAMA & checking wrt GTM  & precision and recall of & 15\\ & & \#$\pre$ and \#$\eff$ & 25 actions\\
         \hline
	\end{tabular}
	\caption{Evaluation of action models (GTM: ground-truth model)}
	\label{table:models_comparison2}
\end{table}

>>>>

This section serves two purposes: first we introduce basic planning concepts as well as the classical planning model that we will use throughout the rest of the paper; and secondly, we summarize the existing approaches to learn classical planning action models and highlight our contributions comparing with related work.

@@

A {\em state} $s$ is a full assignment of values to fluents and we explicitly include negative literals $\neg f$ in states, i.e. $|s|=|F|$, so the size of the state space is $2^{|F|}$. Like in PDDL~\cite{fox2003pddl2}, we assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ such that $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $s=\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$ . A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan for $P$ is {\em optimal} if it has minimum length.

@@

Our approach copes with the partial observability of the plan execution adopting the \emph{open world assumption}. With regard to the observed states in a plan trace, we say that $\tup{s_0, s_1, \ldots, s_n}$ is a fully-observable (\FO) state trajectory if every state $s_i$ is a full assignment of values to fluents and the minimal action sequence to transit from state $s_i$ to state $s_{i+1}$ is composed of a single action; that is, $\theta(s_i,\tup{a})=s_{i+1}$. Otherwise, it is a partially-observable (\PO) state trajectory, meaning that at least one state $s_i$ is a partial assignment of values to fluents in which one or more literals are missing (formally, $|s_i|<|F|$). This implies that in a \PO state trajectory all fluents of a state $s_i$ might be missing, in which case, $s_i$ is a \emph{missing} or \emph{empty} state. This general definition of \PO gives rise to two particular cases:

\begin{itemize}
\item when \textbf{all} the $n-1$ intermediate states of a trajectory $s$ are \textbf{missing}, $s$ is a \emph{non-observable} (\NO) state trajectory.
\item when \textbf{none} of the $n-1$ intermediate states of a trajectory $s$ is \textbf{missing}, we will refer to $s$ as a \POstar state trajectory.
\end{itemize}

@@

\begin{table}[hbt!]
\centering
\begin{tabular}{c|c|c|}
	     & {\bf \# intermediate states} & {\bf state type} \\ \hline
    \FO & $n-1$  & {\small $\forall i, 1 \leq i < n$}  \\  & & $s_i$ is a full assignment \\ \hline
    \multirow{1}{*}{\POstar} & $n-1$ & {\small $\exists i, 1 \leq i < n$}  \\ & & $s_i$ is a partial assignment \\ \hline
    \multirow{1}{*}{\PO} & $\leq n-1$ & {\small $\exists i, 1 \leq i < n$}   \\  & & $s_i$ is a partial assignment \\ \hline
    \NO & 0 & {\small $\forall i, 1 \leq i < n$}  \\  & & $|s_i|=0$  
\end{tabular}
\caption{Classification of state trajectories accordingly to the observed information.}
\label{tab:state_trajectory}
\end{table}

With regard to the observed actions in a plan trace, we say that an action sequence, $\tup{a_1, \ldots, a_n}$ is a \FO action sequence if it contains all the necessary actions to transit every state $s_{i-1}$ to the corresponding successor state $s_{i}$, for each {\small $1\leq i\le n$}. We say that the action sequence is \PO, if at least one of these necessary actions is missing in the sequence and that is \NO, when the sequence of observed actions is empty.

A plan trace $\tau = \langle s_0, a_1, s_1, a_2, s_2, \ldots, a_n, s_n \rangle$ for a planning frame $\Phi=\tup{F,A}$, holds that $a_i\in A$ for every action in $\tau$ and that $s_i \in \mathcal{L}(F)$ for each {\small $1\leq i\le n$}. Plan traces can be classified accordingly to the type of observed state trajectory (\FO, \POstar, \PO or \NO) and action sequence (\FO, \PO or \NO).

@@

The first column of Table \ref{table:models_comparison1} shows the constraints imposed on the input plan traces with regard to observability. Since all approaches except ours deal only with \FO action sequences, constraints are exclusively concerned with the type of state trajectory. This directly affects the complexity of the task, which can be sorted from the least to the most constrained following this order: 1) \NO, 2) \PO, 3) \POstar, and 4) \FO. Note that \PO is less constrained than \POstar because \PO considers the possibility of having some missing state in the trajectory.

The task of learning from less constrained traces subsumes learning from more constrained ones. Consequently, approaches to learning from, for instance traces with \PO state trajectories, will also be able to learn from traces with \POstar state trajectories. All the approaches analyzed in this work accept the more constrained definition of partial observations of intermediate states \POstar, and most of them also allow the sequence of intermediate states to be empty. Exceptionally, a \NO state sequence in \LOCM is a fully-empty trajectory, with neither initial or final state.

@@

Most approaches assume that a set of predicates and a set of action headers are provided alongside the input traces. Others do not explicitly say so but the fact is that predicates and actions headers are easily extractable from the state sequence and action sequence of the input plan traces, respectively. A requirement, however, is that the input plan traces comprise at least a grounded sample of every predicate and operator schema in the domain model.

@@

Table \ref{table:models_comparison2} summarizes the main characteristics of the evaluation of the learned action models based on the type of evaluation method (first column of Table \ref{table:models_comparison2} -- almost all approaches rely on a comparison between the learned model and a {\em Ground-Truth Model}), the metrics used in the comparison (second column of Table \ref{table:models_comparison2}) and the number of tested domains alongside the size of the training dataset (third column of Table \ref{table:models_comparison2}).

@@

\begin{table}
	\small
	\centering
	\begin{tabular}{ l | c | c | c }
		%& \multicolumn{1}{c}{Evaluation task}
        & \textbf{Evaluation method} & \textbf{Metrics} & \textbf{\#tested domains/}   \\
        &   &   & \textbf{training data size} \\
		\hline			
		\multirow{1}{*}{\ARMS} & cross-validation with a test set & error counting of \#$\pre$ satisfaction  & 6  \\
        & of plan traces & and redundancy & 1,600-4,320 actions\\ & & & (160 plan traces) \\
        \hline
         \SLAF &  manual checking wrt GTM &  ---   & 4\\ & & & 1,000 actions\\
         \hline
		\multirow{1}{*}{\LAMP} & checking wrt GTM  & error counting of extra & 4\\ & & and missing \#$\pre$ and \#$\eff$ & 1,300-6,100 actions\\
           & & & (100-200 plan traces) \\
         \hline
         \AMAN & checking wrt GTM &  error counting of extra &  3 \\ & & and missing \#$\pre$ and \#$\eff$ & 40-200 plan traces\\
         \hline
         \NOISTA & checking wrt GTM  & error counting of extra & 5\\ & & and missing \#$\pre$ and \#$\eff$ & 5,000-20,000 actions\\
         \hline
         \CAMA & checking wrt GTM  &  error counting of extra & 3 \\ & & and missing \#$\pre$ and \#$\eff$ & 15-75 plan traces\\
         \hline
		\LOCMtwo & manual checking wrt GTM &  ---  &   --- \\
         \hline
		\FAMA & checking wrt GTM  & precision and recall & 15\\ & &  & 25 actions\\
         \hline
	\end{tabular}
	\caption{Evaluation of action models (GTM: ground-truth model)}
	\label{table:models_comparison2}
\end{table}
